---
title: "BikeShare Case Study"
autor: Varun Srivatsa
date: "26/06/2022"
output:
  pdf_document: default
  html_document: default
---

# Case Study: How does a Bike-Share Navigate Speedy Success?

## Scenario

The director of marketing believes the company's future success depends on maximizing the number of annual memberships. Therefore, my team wants to understand **how casual riders and annual members use Cyclistic bikes differently**. My team will design a new marketing strategy to **convert casual riders into annual members from these insights.**

**The analysis will follow the 6 phases of the Data Analysis Process: Ask, Prepare, Process, Analyze, Share, and Act (APPASA).**

## ASK

### Key tasks

1.  Business task: How can we convert casual riders into an annual members.
2.  Key Stakeholders: Lily Moreno ( director of marketing and my manager ) and Cyclistic executive team.

### A clear statement of the business task

The financial analysts have concluded that annual membership is much more profitable than single-ride and full-day passes from their analysis. So to make people pot for the yearly membership, our marketing campaign should urge the casual riders to convert to annual riders.

As a solution, we should understand why casual riders would convert to a yearly membership? Based on the insights from the above question, we can achieve the maximum required conversion rate from casual to annual riders.

## PREPARE

### Guiding Questions

1.  **Where is your data Located?**

    Data is downloaded from <https://divvy-tripdata.s3.amazonaws.com/index.html> to local system and then uploaded in RStudio cloud where I could use the R programming language for the analysis.

2.  **How is data organized?**

    Data is segregated into quarters from the year 2013 to 2020 till the first quarters of the latter year. Each year having its CSV file. (I will use the data from past 12 months)

3.  **Are there any issues with bias or credibility in this data? Does your data ROCCC?**

    The data has been collected directly from the company's customers, that is, bike riders so there is no issue of bias and credibility for the same reason. It is also Reliable, Original, Comprehensive, Current, and Cited, which satisfies ROCCC.

4.  **How are you accessing licensing, privacy, security, and accessibility?**

    The data was collected by Motivate International Inc. under the following license <https://www.divvybikes.com/data-license-agreement>. Also the data-set does not contain any personal information about its customers (or riders) to violate the privacy.

5.  **How did you verify the data's integrity?**

    The qualities required to verify the data integrity are accuracy, completeness, consistency, and trustworthiness. The data is complete as it contains all the required components to measure the entity. The data is consistent across the years with year having its CSV file which is organized in an equal number of columns and same data types. As the credibility was proven before, it is also trustworthy.

6.  **How does it help to answer your question?**

    By creating new features from existing ones like rideable_type, started_at, and ended_at(which are date-timestamp variables), we can deduce relationship between annual members and casual riders. The relationship analyzed will be useful to answer the question, that is, convert casual riders to annual members.

7.  **Are there any problems with the data?**

    Yes, the data had a couple of problems. There are few rows with "N/A" values which needs to be removed. Also, there are duplicates which have to be eliminated.

Installed required packages.

```{r}
#install.packages(c("tidyverse","ggplot2","lubridate","dplyr","readr","geosphere","scales", "janitor"))
```

Loading the installed packages above into the work-space.

```{r}
library(tidyverse)
library(ggplot2)
library(lubridate)
library(dplyr)
library(readr)
library(geosphere)
library(scales)
library(janitor)
```

Let's load the data stored in CSV file from (June-2021 to May-2022) i.e., 1 years of data

```{r}
tripdata_202106 <- read.csv("./data/202106-divvy-tripdata.csv")
tripdata_202107 <- read.csv("./data/202107-divvy-tripdata.csv")
tripdata_202108 <- read.csv("./data/202108-divvy-tripdata.csv")
tripdata_202109 <- read.csv("./data/202109-divvy-tripdata.csv")
tripdata_202110 <- read.csv("./data/202110-divvy-tripdata.csv")
tripdata_202111 <- read.csv("./data/202111-divvy-tripdata.csv")
tripdata_202112 <- read.csv("./data/202112-divvy-tripdata.csv")
tripdata_202201 <- read.csv("./data/202201-divvy-tripdata.csv")
tripdata_202202 <- read.csv("./data/202202-divvy-tripdata.csv")
tripdata_202203 <- read.csv("./data/202203-divvy-tripdata.csv")
tripdata_202204 <- read.csv("./data/202204-divvy-tripdata.csv")
tripdata_202205 <- read.csv("./data/202205-divvy-tripdata.csv")
```

```{r}
glimpse(tripdata_202106)
```

Create a list with all the data sets and quickly inspect all the data frames

```{r}
data_list = list(tripdata_202106, tripdata_202107, tripdata_202108, tripdata_202109, tripdata_202110, tripdata_202111, tripdata_202112, tripdata_202201, tripdata_202202, tripdata_202203, tripdata_202204, tripdata_202205)
```

Print column names

```{r}
column_names <- colnames(data_list[[1]])
print(column_names)
```

Let's ensure data integrity by checking the column names of all CSV files for consistency. Check that all data frames include the same columns in the same order.

```{r}
for (df in data_list) {
  df_colnames <- colnames(df)
  print(column_names %in% df_colnames)
}
```

### Further Clean Up

Next step is to check the data for errors, inconsistencies, duplicate observations, empty rows or columns, NA values and outliers. In other words, we make sure the data provided is clean and ready for analysis.

Check that all data frames include the same data types for the the same variable across all data sets.

The compare_df_cols function will compare all data-frames and return the mismatched columns.

```{r}
compare_df_cols(data_list, return = c("mismatch"), bind_method = c("bind_rows"))
```

From the above query, we can see that all variables have consistent data types.

Now, let's join the elements in the list into a one single data frame.

```{r}
all_trips <- bind_rows(data_list)
head(all_trips)
```

```{r}
glimpse(all_trips)
```

I see few missing rows in the above glimpse, so lets remove those. First, let's replace blank values with NA.

```{r}
all_trips[all_trips==""] <- NA
glimpse(all_trips)
```

Let's check for missing values in the data frame.

```{r}
sapply(all_trips, function(x) sum(is.na(x)))
```

The datasets contain missing values as listed above. This may cause inaccuracies in the data analysis.

Let's drop all null values in the data frame.

```{r}
all_trips <- drop_na(all_trips)
glimpse(all_trips)
```

Are there any duplicated records?

```{r}
all_trips <- all_trips[!duplicated(all_trips$ride_id),]
```

```{r}
glimpse(all_trips)
```

Seems like there isn't any duplicate values.

Let's get a list with the bike types.

```{r}
bike_types <- list(unique(all_trips$rideable_type))
print(bike_types)
```

Get a list with the unique membership names

```{r}
member_type <- list(unique(all_trips$member_casual))
print(member_type)
```

## PROCESS

### Guiding Questions

1.  **What tools are you choosing and Why?**

    The entries in the trips tables are from the years 2013 to 2020, which is enormous. Since this is the case it is always easy and helpful to navigate through the data using either databases (like SQL) or R programming language. I'll be using R language to deal withe the data in this case study.

2.  **What steps have you taken to ensure that your data is clean?**

    a\. I have concatenated all the CSV files of each year into a single data frame

    b\. Removed all the empty rows and columns from the concatenated data frame.

    c\. Check the unique values in each variable using count() so that there is no miss spelling anywhere.

    d\. Omitted N/A values from the entire data frame.

    e\. Removed duplicates.

3.  **How can you verify that your data is clean and ready to analyze?**

    After performing all the cleaning task mentioned above, I ran the below functions to verify:

    a\. Used filter() to check if there were any missing values.

    b\. Used count() to check the unique values of each variable.

    c\. Used duplicated() to check for any duplicates present.

4.  **Have you documented your cleaning process so you can review and share those results?**

    Yes, please find the below comments and snippets for the documentation.

### Let's get started: Data Manipulation

First, let's convert started_at and ended_at columns from character to timestamp.

```{r}
all_trips$started_at = as.POSIXct(all_trips$started_at, format = "%Y-%m-%d %H:%M:%S")
all_trips$ended_at = as.POSIXct(all_trips$ended_at, format = "%Y-%m-%d %H:%M:%S")
```

Two new columns were added to compute the following.

-   Time spent (in hours) on each trip (column name is labelled as time_difference_hours)

-   Distance traveled (in Kilometers) on each trip (column name is labelled as distance_km)

The new dataframe is labelled as "all_trips_2".

```{r}
all_trips_2 <- mutate(all_trips, time_difference_hours = difftime(ended_at, started_at, units = "hours")) %>% 
  mutate(all_trips, distance_km = distHaversine(cbind(start_lng, start_lat), cbind(end_lng, end_lat))*0.001)
```

```{r}
head(all_trips_2)
```

Perform some calculations to see if the data makes sense

```{r}
summary(all_trips_2$distance_km)
```

Median distance is 1.6km, maximum is way too far, at 1192.24 km. Let's investigate this outlier

```{r}
outlier_df <- filter(all_trips_2, (distance_km > 1000))
print(outlier_df$ride_id)
print(outlier_df$time_difference_hours)
```

1192.24 km in \~3min? Doesn't look right

```{r}
print(paste("Distance (km): ", outlier_df$distance_km, " Duration: ", outlier_df$time_difference_hours, " Start Station: ", outlier_df$start_station_name, " End Station: ", outlier_df$end_station_name))
```

Let's remove the outlier and show the summary again. This new dataframe is used for subsequent data analysis.

```{r}
all_trips_cleaned <- filter(all_trips_2, !(distance_km > 1000 & time_difference_hours < 0.05))
summary(all_trips_cleaned$distance_km)
```

```{r}
summary(all_trips_cleaned)
```

## ANALYSIS

### Scope of analysis

This report will analyse the user trends based on the historical data from June 2021 to May 2022.

### Key Findings

The data analysis revealed the following:

-   Annual members made more trips compared to casual riders. This is expected because the annual members paid a fixed fee for unlimited 45-minutes rides, which may explain the propensity to maximize bike usages.

-   The demand for bike rental increases during the summer period and decreases during the winter period. This observation is similar across casual riders and annual members. One plausible reason for this observation is that people are likely to go out during the summer and less likely to ride a bike during the winter.

-   Among the casual riders, Saturdays and Sundays are the most popular days. However the same was not observed in the annual members; bike usage dropped by around 11.3% from weekday to weekend.

-   For casual riders, the top 5 start and end bike stations are located near attractions. In contrast, for annual members, the top 5 start and end bike stations are located near residential areas.

### Number of rides completed by user type

The annual members (referred to as members in the chart) made more trips than the casual riders.

```{r}
ggplot(all_trips_cleaned, aes(x=member_casual)) +
  geom_bar(fill = "Orange") +
  labs(
    title = "Number of rides completed by user type",
    subtitle = "For the period between June 2021 and May 2022",
    x = "User Type",
    y = "Number of riders (in millions)") +
  scale_y_continuous(labels = label_number(suffix = "M", scale = 1e-6)) +
  geom_text(stat = 'count', aes(label=..count..), vjust=+2, color="white")
```

### Total distance (in kilometers) traveled by user type (please note the assumption and limitation)

The annual members traveled more kilometers than the casual riders. When computing this distance between the starting and ending bike station (by long_lat coordinates), it is assumed that the riders traveled from one point to another without making any detours.

```{r}
data_bar2 <- all_trips_cleaned %>% 
  group_by(member_casual) %>% 
  summarise(distance_km=sum(distance_km, na.rm=TRUE))

ggplot(data_bar2, aes(x=member_casual, y=distance_km)) +
  geom_bar(stat= "identity", fill= "red") +
  labs(
    title = "Distance travelled by user type",
    subtitle = "For the period between June 2021 and May 2022",
    x = "User type",
    y = "Distance travelled (kilometers)") +
  scale_y_continuous(labels = label_number(suffix = " M", scale = 2e-6)) +
  geom_text(aes(label=round(stat(y),2)), vjust=+2, color="white")
```

Although annual members traveled longer total distance, the average distance traveled between annual members and casual riders is roughly the same.

```{r}
data_bar2.1 <- all_trips_cleaned %>% 
  group_by(member_casual) %>% 
  summarise(average_distance_km=mean(distance_km, na.rm=TRUE))
print(data_bar2.1)
```

### Hours cycled by user type

Annual members cycled less hours than casual riders.

```{r}
data_bar3 <- all_trips_cleaned %>% 
  group_by(member_casual) %>% 
  summarise(time=sum(time_difference_hours, na.rm=TRUE))

ggplot(data_bar3, aes(x=member_casual, y=time)) +
  geom_bar(stat = "identity", fill = "pink") +
  labs(
    title = "Hours cycled by user type",
    subtitle = "For the period between June 2021 and May 2022",
    x = "User type",
    y = "Hours cycled") +
  scale_y_continuous(labels = label_number(suffix = " Hrs", scale = 1e-6)) +
  geom_text(aes(label=round(time, 2)), vjust=+2, color="black")
```

On average, annual members cycled 0.3.8 hours less than casual riders.

```{r}
data_bar3.1 <- all_trips_cleaned %>% 
  group_by(member_casual) %>% 
  summarise(average_time=mean(time_difference_hours, na.rm=TRUE))
print(data_bar3.1)
```

### Bike preference by user type

Classic bike is the most preferred bike type among both casual riders and annual members.

```{r}
data_bar4 <- all_trips_cleaned %>% 
  group_by(member_casual, rideable_type) %>% 
  summarise(count_of = n())

ggplot(data_bar4, aes(x=member_casual, y=count_of, fill = rideable_type)) +
   geom_bar(stat="identity") +
  labs(
    title = "Bike preference by user type",
    subtitle = "For the period between June 2021 and May 2022",
    fill = "Bike type",
    x = "User type",
    y = "Number of users") +
  geom_text(aes(label=count_of), position = position_stack(vjust = .5), color="white") +
  scale_y_continuous(labels = label_number(suffix = " M", scale = 1e-6))
```

**Number of rides completed by month by user type**

The **summer period** (around June to September) saw **an increase** in rides completed, while the **winter period** (around Nov to February) saw a **marked reduction** in rides completed. This trend is similar in both user type, likely because people are less likely to go out in the winter.

```{r}
data_bar5 <- mutate(all_trips_cleaned, start_month_year = floor_date(as_date(started_at), "month")) %>%
  group_by(start_month_year, member_casual) %>%
  summarise(count_of = n())

ggplot(data_bar5, aes(x=start_month_year, y=count_of, fill="orange"))+
  geom_bar(stat="identity") +
  facet_wrap(~member_casual)+
  labs(
    title = "Number of rides completed by month by user type",
    subtitle = "For the period between June 2021 and May 2022",
    x = "Month",
    y = "Number of rides completed") +
  geom_text(aes(label=count_of), position = position_stack(vjust = .5), color="black", angle = 90) +
  scale_y_continuous(labels = label_number(suffix = " M", scale = 1e-6)) +
  scale_x_date(date_labels = "%b %y", date_breaks = "2 month") +
  theme(legend.position = "none")
```

**Number of rides completed by day by user type**

Among the casual riders, there is a visible **increase of 73.7%** in bike rentals on **weekends** . This suggests that the casual riders used the bikes for leisure purposes predominantly. This finding corroborates the finding described in the next section.

On the contrary, there is a slight **decrease of 11.3%** in bike rentals on **weekends** among the annual members. This suggests that the annual members used the bikes for non-leisure or work purposes predominantly. This finding corroborates the finding described in the next section.

```{r}
data_bar6 <- mutate(all_trips_cleaned, start_day = weekdays(started_at)) %>%
  group_by(start_day, member_casual) %>%
  summarise(count_of = n())

level_order <-c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday', 'Sunday')

ggplot(data_bar6, aes(x=factor(start_day, level = level_order), y=count_of, fill=member_casual))+
  geom_bar(stat="identity") +
  facet_wrap(~member_casual) +
  labs(
    title = "Number of rides completed by day by user type",
    subtitle = "For the period between June 2021 and May 2022",
    x = "Day",
    y = "Number of rides completed",
    fill = "User type") +
  geom_text(aes(label=count_of), position = position_stack(vjust = .5), color="black", angle = 90) +
  scale_y_continuous(labels = label_number(suffix = " M", scale = 1e-6)) +
  theme(axis.text.x=element_text(angle = 90))
```

**Top 5 start stations by user types**

For the casual riders, the top five start stations are largely located near **places of attractions** . This suggests that the casual riders rented bikes to tour around the attractions.

For annual members, the top five start stations are largely located near **residential areas** . This suggests that the annual members rented bikes for non-leisure or work purposes.

```{r}
table1 <- all_trips_cleaned %>%
  group_by(member_casual, start_station_name) %>%
  summarise(count_of=n()) %>%
  arrange(desc(count_of)) %>%
  na.omit(start_station_name)

  ## Table 1.1 - By casual riders ##
table1.1 <- filter(table1, member_casual =="casual") %>%
  rename(number_of_trips = count_of) %>%
  slice(1:5)

  #Table 1.2 - By members ##
table1.2 <- filter(table1, member_casual =="member") %>%
  rename(number_of_trips = count_of) %>%
  slice(1:5)

print(table1.1)
```

```{r}
print(table1.2)
```

**Top 5 end stations by user types**

The top five end stations are **similar** to the top five start stations as described earlier.

```{r}
table2 <- all_trips_cleaned %>%
  group_by(member_casual, end_station_name) %>%
  summarise(count_of=n()) %>%
  arrange(desc(count_of)) %>%
  na.omit(end_station_name)

  ## Table 2.1 - By casual riders ##
table2.1 <- filter(table2, member_casual =="casual") %>%
  rename(number_of_trips = count_of) %>%
  slice(1:5)

  #Table 2.2 - By members ##
table2.2 <- filter(table2, member_casual =="member") %>%
  rename(number_of_trips = count_of) %>%
  slice(1:5)

print(table2.1)
```

```{r}
print(table2.2)
```

## SHARE

**Recommendations**

Three key recommendations are proposed to convert casual riders to annual members:

-   Identify casual riders who typically start and end their ride sessions near residential areas and offer incentives (e.g. discounts, lucky draws, etc) to convert them into annual members.

-   Offer an annual membership for weekday rental to increase number of annual members.

-   Conduct a market survey with the casual riders, asking if they would purchase an annual package if there is a bike station near their residence. If the response is positive, consider identifying key residential areas to build new bike stations.

Two other recommendations are proposed to increase sales:

-   Partner with tour operators to offer a combination package (e.g. attraction pass + bike pass) to increase number of casual riders.

-   Increase marketing campaigns during the summer season.

**Further exploration**

-   Collect data to assess whether there are sufficient bikes to cater to the demand in the top five bike stations

-   Conduct marketing promotions and collect price vs demand data to determine the appropriate price point.

**Back to the three questions:**

**1. How do annual members and casual riders use Cyclistic bikes differently?**

Casual users prefer bike rides during the weekends, and mostly for leisure. They ride for a longer time and slightly longer distances than annual members. Annual members mostly use bikes for commutes. The use of bike rides peaks during summer months.

**2. Why would casual riders buy Cyclistic annual memberships?**

Casual riders would buy the annual memberships if there was a discount for longer rides or for weekends, or seasonal (summer) discount.

**3. How can Cyclistic use digital media to influence casual riders to become members?**

Cyclistic can create ads on social media in the areas surrounding the most touristic places (most popular destinations). It could also collaborate with touristic sites to provide discount or free entrance tickets to casual users converting into annual members.
